---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# GEAtlas

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check-bioc](https://github.com/almeidasilvaf/GEAtlas/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/almeidasilvaf/GEAtlas/actions)
[![Codecov test coverage](https://codecov.io/gh/almeidasilvaf/GEAtlas/branch/main/graph/badge.svg)](https://codecov.io/gh/almeidasilvaf/GEAtlas?branch=main)
<!-- badges: end -->

The goal of `GEAtlas` is to download RNA-seq data from NCBI SRA,
preprocess them, map to the reference genome, and quantify the expression
at the gene level. The goal of GEAtlas is to make RNA-seq data analysis
pipelines reproducible, with a framework built on state-of-the art methods
and softwares.

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `GEAtlas` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("GEAtlas")
```

And the development version from [GitHub](https://github.com/almeidasilvaf/GEAtlas) with:

```{r 'install_dev', eval = FALSE}
remotes::install_github("almeidasilvaf/GEAtlas")
```

## Code of Conduct

Please note that the `GEAtlas` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

## Development tools

* Continuous code testing is possible thanks to [GitHub actions](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/)  through `r BiocStyle::CRANpkg('usethis')`, `r BiocStyle::CRANpkg('remotes')`, and `r BiocStyle::CRANpkg('rcmdcheck')` customized to use [Bioconductor's docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh) and `r BiocStyle::CRANpkg('covr')`.
* The [documentation website](http://almeidasilvaf.github.io/GEAtlas) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
* The code is styled automatically thanks to `r BiocStyle::CRANpkg('styler')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.

For more details, check the `dev` directory.

This package was developed using `r BiocStyle::Biocpkg('biocthis')`.


